Sequoidal Categories and Transfinite Games: Towards a Coalgebraic Approach to Linear Logic\footnote{This work was partially supported by [TODO: find the appropriate grant codes]}

### 2017-04-08 round-corner/paper/lipics-v2016-authors/sequoidal.tex
Draft of paper.

%% Other macros
\newcommand*\from{\colon}
\def \inv {^{-1}}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\pr}{pr}
\newcommand{\tensor}{\otimes}
\newcommand{\sequoid}{\oslash}
\renewcommand{\implies}{\multimap}
\newcommand{\comp}[2]{#1 \circ #2}
\newcommand{\cprd}{\sqcup}
\newcommand{\bigcprd}{\bigsqcup}
\newcommand{\C}{\mathcal C}
\newcommand{\F}{\mathcal F}
\newcommand{\G}{\mathcal G}
\newcommand{\suchthat}{\;\colon\;}
\newcommand{\varsuchthat}{\;\mid\;}
\newcommand{\OP}{\{O,P\}}
\newcommand{\emptyplay}{\epsilon}
\newcommand{\prefix}{\sqsubseteq}
\newcommand{\pprefix}{\sqsubsetneqq}
\newcommand{\ppprefix}{\sqsubset}
\DeclareMathOperator{\assoc}{assoc}
\DeclareMathOperator{\lunit}{lunit}
\DeclareMathOperator{\runit}{runit}
\DeclareMathOperator{\sym}{sym}
\newcommand{\braid}{\sym}
\newcommand{\blank}{\;\underline{\hspace{1.5ex}}\;}
\newcommand{\der}{{\mathtt{der}}}
\newcommand{\mult}{{\mathtt{mult}}}
\DeclareMathOperator{\wk}{wk}
\newcommand{\toisom}{{\xrightarrow{\cong}}}
\newcommand{\passoc}{{\mathtt{passoc}}}
\newcommand{\pcomm}{{\mathtt{pcomm}}}
\newcommand{\run}{{\mathtt{r}}}
\newcommand{\lun}{{\mathtt{l}}}
\newcommand{\fcoal}[1]{{\leftmoon #1 \rightmoon}}
\renewcommand{\subset}{\subseteq}
\let\oldemptyset\emptyset
\let\emptyset\varnothing
\newcommand{\s}{\mathfrak{s}}
\newcommand{\dist}{{\mathtt{dist}}}
\newcommand{\dec}{{\mathtt{dec}}}
\renewcommand{\int}{{\mathtt{coh}}}
\renewcommand{\coh}{\int}
\DeclareMathOperator{\CCom}{CCom}
\newcommand{\catname}[1]{{\normalfont\textbf{#1}}}
\newcommand{\Set}{\catname{Set}}
\newcommand{\Ord}{\catname{Ord}}
\newcommand\oppcat[1]{#1^{\mathrm{op}}}
\newcommand{\bN}{{\mathbb{N}}}
\newcommand{\OK}{{\mathtt{OK}}}
\newcommand{\zero}{{\mathtt{zero}}}
\renewcommand{\succ}{{\mathtt{succ}}}
\newcommand{\cell}{{\mathtt{cell}}}
\newcommand{\Var}{{\mathtt{Var}}}
\renewcommand{\read}{{\mathtt{read}}}
\newcommand{\wwrite}{{\mathtt{write}}}

@@

\begin{abstract}
  In \cite{laird02}, Laird introduces the concept of a \emph{sequoidal category} as a formalization of causality in game semantics.  A sequoidal category is like a monoidal category with an extra connective, $\sequoid$, that allows one to construct an exponential object as a final coalgebra \cite{martinsthesis},  Under a further hypothesis, it is possible to show that this final coalgebra allows one to construct cofree commutative comonoids in the category, giving us a model of the exponential connective $\oc$ from linear logic.  In the first part of this note, we review the coalgebraic arguments for constructing the cofree commutative comonoid, which are known but do not yet appear in print.  In the second part, we show that the extra hypotheses are necessary by outlining a definition of \emph{transfinite game}, in which the carriers of the sequoidal coalgebra and the cofree commutative comonoid do not coincide.
\end{abstract}

@@

\begin{definition}
  A \emph{sequoidal category} consists of the following data:
  \begin{itemize}
    \item A symmetric monoidal category $\C$ with monoidal product $\tensor$ and tensor unit $I$, associators $\assoc_{A,B,C}\from(A\tensor B)\tensor C\toisom A\tensor(B\tensor C)$, unitors $\runit_A\from A\tensor I\toisom A$ and $\lunit_A\from I\tensor A\toisom A$ and braiding $\sym_{A,B}\from A\tensor B\to B\tensor A$.
    \item A category $\C_s$
    \item A right monoidal category action of $\C$ on the category $\C_s$.  That is, a functor
      \[
        \blank\sequoid\blank\from\C_s\times\C\to\C_s
        \]
      together natural isomorphisms
      \[
        \passoc_{A,B,C}\from A\sequoid(B\tensor C)\to (A\sequoid B)\sequoid C
        \]
      and
      \[
        \run_A\from A\sequoid I\toisom A
        \]
      subject to the following coherence conditions:
      \begin{equation*}
        \begin{tikzcd}
          A\sequoid(B\tensor(C\tensor D)) \arrow[r, "\passoc_{A,B,C\tensor D}" yshift=0.3em] \arrow[d, "\id_A\sequoid \assoc_{B,C,D}"']
            & (A \sequoid B) \sequoid (C\tensor D) \arrow[r, "\passoc_{A\sequoid B,C,D}" yshift=0.3em]
              & ((A\sequoid B)\sequoid C) \sequoid D \\
          A\sequoid((B\tensor C)\tensor D) \arrow[r, "\passoc_{A,B\tensor C,D}"' yshift=-0.3em]
            & (A\sequoid (B\tensor C)) \sequoid D \arrow[ur, "\passoc_{A,B,C}\sequoid\id_D"']
              &
        \end{tikzcd}
      \end{equation*}
      \begin{equation*}
        \begin{tikzcd}
          A \sequoid (I\tensor B) \arrow[r, "\passoc_{A,I,B}" yshift=0.3em] \arrow[d, "\id_A\sequoid\lunit_B"']
            & (A\sequoid I)\sequoid B \arrow[dl, "\run_A\sequoid\id_B"] \\
          A \sequoid B
            &
        \end{tikzcd}
        \quad
        \begin{tikzcd}
          A \sequoid (B\tensor I) \arrow[r, "\passoc_{A,B,I}" yshift=0.3em] \arrow[d, "\id_A\sequoid\runit_B"']
            & (A\sequoid B) \sequoid I \arrow[dl, "\run_{A\sequoid B}"] \\
          A\sequoid B
            &
        \end{tikzcd}
      \end{equation*}

@@

\begin{definition}
  Let $\C=(\C,\C_s,J,\wk)$ be a sequoidal category.  We say that $\C$ is an \emph{inclusive sequoidal category} if $\C_s$ is a full-on-objects subcategoryof $\C$ containing the monoidal isomorphisms and the morphisms $\wk_{A,B}$, $J$ is the inclusion functor and $J$ reflects isomorphisms.

  If $\C$ is an inclusive sequoidal category, we say that $\C$ is \emph{Cartesian} if $\C_s$ has all products and these are preserved by $J$.  In that case, we say that $\C$ is \emph{decomposable} if the natural transformations
  \begin{gather*}
    \dec_{A,B} = \langle \wk_{A,B}, \comp{\wk_{A,B}}{\sym_{A,B}}\rangle\from A\tensor B\to (A\sequoid B)\times (B\sequoid A) \\
    \dec^0 \from I \to 1
  \end{gather*}
  are isomorphisms and we say that $\C$ is \emph{distributive} if the natural transformations
  \begin{gather*}
    \dist_{A,B,C} = \langle \pr_1\sequoid \id_C,\pr_2\sequoid\id_C\rangle\from (A\times B)\sequoid C\to (A\sequoid C)\times (B\sequoid C) \\
    \dist_{(A_i\colon i\in I),C} = \langle (\pr_i\sequoid\id_C\suchthat i\in I)\rangle \from \left(\prod_{i\in I} A_i\right) \sequoid C \to \prod_{i\in I}(A_i\sequoid C) \\
    \dist_{A,0}\from 1\sequoid A\to 1
  \end{gather*}
  are isomorphisms.
\end{definition}

We have one further piece of structure available to us:

@@

  The sequoidal exponential $\oc A$, together with the obvious copycat strategy $\alpha\from\oc A\to A\sequoid\oc A$, is the final coalgebra for the endofunctor $A\sequoid\blank$.  In other words, if $B$ is a negative game and $\sigma\from B\to A\sequoid B$ is a morphism then there is a unique morphism $\fcoal{\sigma}\from B\to \oc A$ such that the following diagram commutes:
  \[
    \begin{tikzcd}
      B \arrow[r, "\sigma"] \arrow[d, "\fcoal{\sigma}"']
        & A \sequoid B \arrow[d, "\id_A\sequoid\fcoal{\sigma}"] \\
      \oc A \arrow[r, "\alpha"']
        & A \sequoid \oc A
    \end{tikzcd}
    \]
  We call $\fcoal{\sigma}$ the \emph{anamorphism} of $\sigma$.
\end{proposition}
\begin{proof}
  See \cite{martinsthesis}.  We shall shortly give a proof in the more general case.
\end{proof}

>>>>

%% Other macros
\newcommand*\from{\colon}
\def \inv {^{-1}}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\pr}{pr}
\newcommand{\tensor}{\otimes}
\newcommand{\sequoid}{\oslash}
\renewcommand{\implies}{\multimap}
\newcommand{\comp}[2]{#1 \circ #2}
\newcommand{\cprd}{\sqcup}
\newcommand{\bigcprd}{\bigsqcup}
\newcommand{\C}{\mathcal C}
\newcommand{\F}{\mathcal F}
\newcommand{\G}{\mathcal G}
\newcommand{\W}{\mathcal W}
\newcommand{\suchthat}{\;\colon\;}
\newcommand{\varsuchthat}{\;\mid\;}
\newcommand{\OP}{\{O,P\}}
\newcommand{\emptyplay}{\epsilon}
\newcommand{\prefix}{\sqsubseteq}
\newcommand{\pprefix}{\sqsubsetneqq}
\newcommand{\ppprefix}{\sqsubset}
\DeclareMathOperator{\assoc}{assoc}
\DeclareMathOperator{\lunit}{lunit}
\DeclareMathOperator{\runit}{runit}
\DeclareMathOperator{\sym}{sym}
\newcommand{\braid}{\sym}
\newcommand{\blank}{\;\underline{\hspace{1.5ex}}\;}
\newcommand{\der}{{\mathtt{der}}}
\newcommand{\mult}{{\mathtt{mult}}}
\DeclareMathOperator{\wk}{wk}
\newcommand{\toisom}{{\xrightarrow{\cong}}}
\newcommand{\passoc}{{\mathtt{passoc}}}
\newcommand{\pcomm}{{\mathtt{pcomm}}}
\newcommand{\run}{{\mathtt{r}}}
\newcommand{\lun}{{\mathtt{l}}}
\newcommand{\fcoal}[1]{{\leftmoon #1 \rightmoon}}
\renewcommand{\subset}{\subseteq}
\let\oldemptyset\emptyset
\let\emptyset\varnothing
\newcommand{\s}{\mathfrak{s}}
\newcommand{\dist}{{\mathtt{dist}}}
\newcommand{\dec}{{\mathtt{dec}}}
\renewcommand{\int}{{\mathtt{coh}}}
\renewcommand{\coh}{\int}
\DeclareMathOperator{\CCom}{CCom}
\newcommand{\catname}[1]{{\mathsf{#1}}}
\newcommand{\Set}{\catname{Set}}
\newcommand{\Ord}{\catname{Ord}}
\newcommand\oppcat[1]{#1^{\mathrm{op}}}
\newcommand{\bN}{{\mathbb{N}}}
\newcommand{\OK}{{\mathtt{OK}}}
\newcommand{\zero}{{\mathtt{zero}}}
\renewcommand{\succ}{{\mathtt{succ}}}
\newcommand{\Var}{{\mathtt{Var}}}
\renewcommand{\read}{{\mathtt{read}}}
\newcommand{\wwrite}{{\mathtt{write}}}
\newcommand{\Rel}{\catname{Rel}}
\newcommand{\inr}{{\mathsf{inr}}}
\newcommand{\inl}{{\mathsf{inl}}}
\newcommand{{\Na}}{\bN}
\newcommand{{\cell}}{{\mathsf{cell}}}
\newcommand{\fix}{{\mathsf{fix}}}
\newcommand{\eq}{{\mathsf{eq}}}
\newcommand{\co}{{\textrm{\textexclamdown}}}

@@

\begin{abstract}
The non-commutative sequoid operator $\sequoid$ on games was introduced by one of the authors in order to capture algebraically the presence of state in history-sensitive strategies in game semantics.  It imposes a causality relation on to the  tensor product of games, so that coalgebras for the  functor  $A\sequoid\blank$  - i.e. morphisms from $S$ to $A \sequoid S$ may be viewed as state transformers.  If $A\sequoid\blank$ has a final coalgebra, $\oc A$, then the anamorphism of such a state transformer encapsulates its explicit state, so that it is shared only between successive invocations.

We shall study  conditions under which such a final coalgebra $\oc A$ for the functor $A\sequoid\blank$ is the carrier of a cofree commutative comonoid on $A$. That is, it is a model of the exponential of linear logic, but one in which we can construct imperative objects such as reference cells and stacks coalgebraically, and reason about them coinductively, in a game semantics setting.

We shall review the definition of a sequoidal category, a category-theoretic formalization of the sequoid operator, and the notion of sequoidal decomposition of a symmetric monoidal category.  We show that given such a decomposition, the final coalgebra !A may be endowed with the structure of the cofree commutative comonoid if it is the bifree algebra for $A\sequoid\blank$, or if it is a limit of sequoidal powers (following the construction  of the cofree commutative comonoid by Melli\'es, Tabareau and Tasson), or if the functor $\oc$ is strong monoidal - for all  $A$ and $B$, the objects $\oc (A \times B)$ and $\oc A \tensor \oc B$ are naturally isomorphic.

We establish necessity of this final condition  by giving an example of a sequoidally decomposable category of games in which plays will be allowed to have transfinite length. In this category, the final coalgebra for the functor $A\sequoid\blank$ is not the cofree commutative comonoid over $A$: we illustrate this by explicitly contrasting the final sequence for the functor $A\sequoid\blank$ with the chain of sequoidal powers used in the Melli\'es-Tabareau-Tasson construction.
\end{abstract}

\section{Introduction}

Games models of pure functional programming languages such as PCF (\cite{ajmPcf} and \cite{hoPcf}, for example) tend to include the requirement that strategies be \emph{history-free}: that is, the response to a particular position may depend only upon the last move played, and not upon what has gone before.  This makes sense for stateless languages, where the program should have no `memory' to recall what has gone before.

If we include history-sensitive strategies in our models, then we can start to recover some stateful behaviour (see \cite{SamsonGuyIAPassive}, for example), since history-sensitive strategies have a built in state.  This concept of state is often hard to work with: since the representation as a strategy is so intensional, there is no object representing the state, and the state of the program must be inferred from the behaviour of the strategy.

At the other extreme, we have traditional, domain-theoretic semantics, in which stateful programs are represented as \emph{state transformers} -- that is, functions $W\times X \to W\times Y$, where $W$ is an object representing the state.

State transformers are typically fairly straightforward to reason about, and we would like to have a similar technique available in our games models.  This gap was bridged by one of the authors in \cite{laird02} with the introduction of the new \emph{sequoid} operator $\blank\sequoid\blank$ on games, which the same as the tensor product $\blank\tensor\blank$, except that play is required to start in the left hand game.  We interpret a state transformer as a coalgebra for the functor $A\sequoid\blank$ -- that is, a morphism from $S$ to $A\sequoid S$, so that the state $S$ is only modified after the program $A$ has been consulted.  The benefit of this explicit ordering is that we may model multiple iterations of the state transformer as coalgebras:
\[
  S \to A \sequoid (A \sequoid \cdots \sequoid (A \sequoid S)\cdots)
  \]
where the copies of $A$ are started in order from left to right.  The state from a previous invocation of $A$ is passed on to the next, until the final modified state finally appears at the right.

Now we introduce the key part of the construction: suppose that the functor $A\sequoid\blank$ has a final coalgebra $\oc A\to A\sequoid \oc A$.  Then a coalgebra $S\to A\sequoid S$ (i.e., a state transformer) naturally gives rise to a morphism $S \to \oc A$ in which the state is completely encapsulated and we are allowed to invoke the program $A$ as many times as we like.  We will show that we can construct various history-sensitive strategies denoting stateful programs using precisely this strategy; in particular, we will give a coalgebraic construction of the $\cell$ strategy from \cite{SamsonGuyIAPassive}, which is the key to their model for Idealized Algol.

Even more intriguingly, the game $\oc A$ that arises as the final coalgebra for the functor $A\sequoid\blank$ has the structure of an exponential object in the category of games: it is the cofree commutative comonoid over $A$.  We will spend the rest of the paper examining why this is, and what general conditions we need in order to show that the final coalgebra for $\oc A$ and the cofree commutative comonoid over $A$ coincide.  For this, we will need the abstract definition of a \emph{sequoidal category}, which is an extension of a monoidal category in order to accommodate the sequoid operator.  We will first consider the explicit formula for the cofree commutative comonoid given by Melli\'es, Tabareau and Tasson in \cite{MelliesCofCommCom} and show that it can be given a sequoidal interpretation.  The Melli\'es-Tabareau-Tasson construction works for finite games, but its hypotheses do not hold in the case of the category of games with infinitary winning conditions and winning strategies, even though the final coalgebra for $A\sequoid\blank$ and the cofree commutative comonoid over $A$ still coincide in these categories.

Instead, we will show that if our sequoidal category has all products then the cofree commutative comonoid structure on $\oc A$ may be constructed directly from the sequoidal structure and the definition of $A\sequoid\blank$ as a final coalgebra for $A\sequoid\blank$, provided that we insist on a further hypothesis: that the functor $\oc$ is monoidal from the Cartesian category to the monoidal category: i.e., if we have an isomorphism between $\oc A \tensor \oc B$ and $\oc(A\times B)$ for all $A,B$.  We show that this hypothesis is automatically satisfied whenever $\oc A$ has the structure of a \emph{bifree algebra} for $A\sequoid\blank$; i.e., when it simultaneously has the structure of final coalgebra and initial algebra.  However, the exponential in the category of win-games is not an initial algebra for $A\sequoid\blank$ \cite{martinsthesis}.

Lastly, we will provide an illuminating counterexample: a category of games in which the final coalgebra for $A\sequoid\blank$ and the cofree commutative comonoid over $A$ do not coincide.  This is the category of games where plays are allowed to have transfinite length.  We will show that the Melli\'es-Tabareau-Tasson construction fails badly in this setting, but that we may still construct the final coalgebra for $A\sequoid\blank$ as the limit of its final sequence.  In this setting, the final coalgebra is the game where we have `only' $\omega$-many copies of the sub-game to play in, whereas in the cofree commutative comonoid we may open transfinite sequences of the subgame.

\section{Defining Higher-order Stateful Objects, Coalgebraically}
In this section we motivate the study of the coalgebraically derived cofree comonoid in game semantics by considering a similar but simpler and more familiar phenomenon. A \emph{state-transformer} in a symmetric monoidal category is a morphism $f:A \otimes S \rightarrow B \otimes S$ taking an argument together with an input state to a result together with a input state. A well-studied \cite{jacobs} technique in semantics is to  use an appropriate final coalgebra to \emph{encapsulate} the state in such a transformer, allowing multiple  successive invocations, each of which passes its output state as an input state to the next invocation.

For example, consider the category $\Rel$ of sets and relations, with symmetric monoidal structure given by the cartesian product (with unit $I$, the singleton set $\{*\}$). This has finite (bi)products given by the disjoint union of sets: define the functor $F(A,S) = (A \otimes S) \oplus I$

For any object (set) $A$, let $A^*$ be the set $A^*$ of finite sequences of elements of $A$ (i.e. the carrier of the free monoid on $A$), and $\alpha:A^* \rightarrow F(A,A^*)$ be the morphism $\{(\varepsilon,\inr(*)\}\cup  \{(aw,(\inl(a,w))\ |\ a \in A,w \in A^*\}$   
\begin{lemma}$(A^*,\alpha_A)$ is a final coalgebra  for $F(A,\_)$.
\end{lemma}
\begin{proof}$\Rel$ may be cpo-enriched with the inclusion order: $A^*$ is a \emph{minimal invariant} for $F(A,\_)$ with respect to this order.
\end{proof}
Since we have a natural transformation (left injection) $\inl_{A,S}:A \otimes S \rightarrow F(A,S)$, we may encapsulate the state in the  state transformer $f:S \rightarrow A \otimes S$ by taking the \emph{anamorphism} of $f;\inl_{A,B}:S \rightarrow F(A,S)$, --- i.e. the unique $F(A,\_)$-coalgebra morphism from $(S,f;\inl)$ into $(A^*,\alpha_{A})$. This  is  a morphism from an initial state $S$ into $A^*$: by definition, composing it with $\alpha:A \rightarrow F(A,A)$ (which we can think of as  \emph{invoking} our stateful object) returns a copy of $f$ and uses it to update the internal state.

By distributivity of $\oplus$ over  $\otimes$ we have a natural transformation $t: F(A,S) \otimes S' \rightarrow F(A,S \otimes S')$. We may use this to encapsulate  state transformers of general form: given $f:A \otimes S \rightarrow B \otimes S$, taking the anamorphism of the $F(B,\_)$-colagebra:    
$$A^* \otimes S  \stackrel{\alpha_A \otimes S}{\longrightarrow} F(A,A^*) \otimes S \stackrel{t}{\longrightarrow} F(A,A^* \otimes S)$$

 Distributivity of $\oplus$ over $\otimes$ also implies that $F(A \oplus A',S) \cong F(A,S) \oplus F(A',S)$.  This allows state transformers to be aggregated, to construct stateful objects compounded of a series of methods which share access to a common state.
i.e. given morphisms $f_1:A_1 \otimes S \rightarrow B_1 \otimes S, \ldots , f_n:A_n \otimes S \rightarrow B_n \otimes S$, encapsulating  $\dist;(f_1 \oplus \ldots \oplus f_n);\dist^{-1}$, gives a morphism from $(A_1 \oplus \ldots \oplus A_n)^* \otimes S$ into $(B_1 \oplus \ldots \oplus B_n)^*$.

For example, we may represent a reference cell storing integer values as a state transformer $\cell:\Na \rightarrow (\Na \oplus \Na) \otimes \Na$, obtained by aggregating two ``methods'' which share access to a state consisting of a single integer, representing the contents of the cell --- returning a ``read'' of the input state (and leaving it unchanged) or accepting a ``write'' of a new value and using it to update the state.   Thus (with appropriate tagging) it is the relation $\{(i, (\read(i),i))\ |\ i \in \Na\} \cup \{(i,{\mathtt{write}}(j),j)\}$.
The anamorphism of the coalgebra $\cell;\inl:\Na \rightarrow F(\Na \oplus \Na,\Na) $  is the relation from $\Na$ to $(\Na \oplus \Na)^*$ consisting of pairs of the form $(i_1,\read(i_1)^*{\mathtt{write}}(i_2)\read(i_2)^*\ldots)$ Composition with this morphism is precisely the denotation of new variable declaration in the semantics of  Reynolds \emph{Syntactic Control of Interference} (SCI) in $\Rel$ given in \cite{Mcsci}.

Coalgebraic methods thus give us a recipe for constructing and using categorical definitions of stateful semantic objects, avoiding direct definitions which are rather combinatorial to work with.  In order to fully exploit these, however, we may endow $A^*$ with the structure of a \emph{comonoid} by defining morphisms $\delta_A:A^* \rightarrow A^* \otimes A^* = \{(u\cdot v,(u,v))\ |\ u,v \in A^*\}$ and $\epsilon:A \rightarrow I = \{(\varepsilon,*)\}$

 In fact, this is the \emph{cofree comonoid} on $A$ --- there is a morphism $\eta_A:A^* \rightarrow A = \{(a,a)\ | \ a \in A\}$ such that for any comonoid $(B,\delta_B,\epsilon_B)$, composition with $\eta_A$ defines an equivalence (natural in $B$)  between the morphisms from $B$ into $A$, and the comonoid morphisms from $(B,\delta_B,\epsilon_B)$ into  $(A^*,\delta_A,\epsilon_A)$.   
\begin{proposition}$(A^*,\delta,\epsilon)$ is the cofree comonoid on $\Rel$.
\end{proposition}
The definitions of $\delta$ and $\epsilon$, and the proof that this is the cofree comonoid  may be derived from the fact that $(A^*,\alpha_A)$ is a \emph{bifree algebra} for $F(\_, A)$  --- i.e. $(A^*,\alpha^{-1})$ is an initial algebra for $F(A,\_)$ ($\alpha$ must be an isomorphism by Lambek's lemma). We leave this as an exercise.

This structure can be used to interpret procedures which share access to a stateful resource such as a reference cell, the creation of multiple copies of such procedures ($(\_)^*$ is a monoidal comonad)  etc.      
Its main limitation is that we have not defined a \emph{commutative} monoid --- evidently it is not the case that $\delta_A;\sym_{A,A} = \delta_A$ (where $\sym_{A,B}:A \otimes B \rightarrow B \otimes A$ is the symmetry isomorphism for the tensor) for any non-empty set $A$. Thus we can only model procedures with shared access to the same stateful object if the order in which they are permitted to access it is predetermined. This is precisely the situation in SCI, where the typing system allows sharing between across sequential composition, but not between functions and their arguments. In order to model sharing of state between functions without this constraint (and build a Cartesian closed category), we need to endow our final coalgebra with the structure of a cofree \emph{commutative} comonoid.  
The category of sets and relations  does not allow this (the cofree commutative comonoid on an object $A$ in $\Rel$ is given by the set of finite multisets of $A$, which is not a final coalgebra). By properly reflecting the interleaving between function calls, game semantics does support a coalgebraic cofree commutative comonoid, leading to models of stateful higher-order languages such as Idealized Algol and core ML. As we shall show by example and counterexample, the richer structure of games also means that the relationship between initial and final coalgebras, and cofree objects is more subtle  (note, for example, that any final coalgebra in $\Rel$ must be bifree, by its self-duality).

@@

\begin{definition}
  A \emph{sequoidal category} consists of the following data:
  \begin{itemize}
    \item A symmetric monoidal category $\C$ with monoidal product $\tensor$ and tensor unit $I$, associators $\assoc_{A,B,C}\from(A\tensor B)\tensor C\toisom A\tensor(B\tensor C)$, unitors $\runit_A\from A\tensor I\toisom A$ and $\lunit_A\from I\tensor A\toisom A$ and braiding $\sym_{A,B}\from A\tensor B\to B\tensor A$.
    \item A category $\C_s$
    \item A right monoidal category action of $\C$ on the category $\C_s$.  That is, a functor
      \[
        \blank\sequoid\blank\from\C_s\times\C\to\C_s
        \]
      together with natural isomorphisms
      \[
        \passoc_{A,B,C}\from A\sequoid(B\tensor C)\to (A\sequoid B)\sequoid C
        \]
      and
      \[
        \run_A\from A\sequoid I\toisom A
        \]
      subject to the following coherence conditions:
      \begin{equation*}
        \begin{tikzcd}
          A\sequoid(B\tensor(C\tensor D)) \arrow[r, "\passoc_{A,B,C\tensor D}" yshift=0.3em] \arrow[d, "\id_A\sequoid \assoc_{B,C,D}"']
            & (A \sequoid B) \sequoid (C\tensor D) \arrow[r, "\passoc_{A\sequoid B,C,D}" yshift=0.3em]
              & ((A\sequoid B)\sequoid C) \sequoid D \\
          A\sequoid((B\tensor C)\tensor D) \arrow[r, "\passoc_{A,B\tensor C,D}"' yshift=-0.3em]
            & (A\sequoid (B\tensor C)) \sequoid D \arrow[ur, "\passoc_{A,B,C}\sequoid\id_D"']
              &
        \end{tikzcd}
      \end{equation*}
      \begin{equation*}
        \begin{tikzcd}
          A \sequoid (I\tensor B) \arrow[r, "\passoc_{A,I,B}" yshift=0.3em] \arrow[d, "\id_A\sequoid\lunit_B"']
            & (A\sequoid I)\sequoid B \arrow[dl, "\run_A\sequoid\id_B"] \\
          A \sequoid B
            &
        \end{tikzcd}
        \quad
        \begin{tikzcd}
          A \sequoid (B\tensor I) \arrow[r, "\passoc_{A,B,I}" yshift=0.3em] \arrow[d, "\id_A\sequoid\runit_B"']
            & (A\sequoid B) \sequoid I \arrow[dl, "\run_{A\sequoid B}"] \\
          A\sequoid B
            &
        \end{tikzcd}
      \end{equation*}

@@

\begin{definition}
  Let $\C=(\C,\C_s,J,\wk)$ be a sequoidal category.  We say that $\C$ is an \emph{inclusive sequoidal category} if $\C_s$ is a full-on-objects subcategoryof $\C$ containing all isomorphisms in $\C$ and the morphisms $\wk_{A,B}$ and $J$ is the inclusion functor.

  If $\C$ is an inclusive sequoidal category, we say that $\C$ is \emph{Cartesian} if $\C_s$ has all products and these are preserved by $J$.  In that case, we say that $\C$ is \emph{decomposable} if the natural transformations
  \begin{gather*}
    \dec_{A,B} = \langle \wk_{A,B}, \comp{\wk_{A,B}}{\sym_{A,B}}\rangle\from A\tensor B\to (A\sequoid B)\times (B\sequoid A) \\
    \dec^0 \from I \to 1
  \end{gather*}
  are isomorphisms and we say that $\C$ is \emph{distributive} if the natural transformations
  \begin{gather*}
    \dist_{A,B,C} = \langle \pr_1\sequoid \id_C,\pr_2\sequoid\id_C\rangle\from (A\times B)\sequoid C\to (A\sequoid C)\times (B\sequoid C) \\
    \dist_{(A_i\colon i\in I),C} = \langle (\pr_i\sequoid\id_C\suchthat i\in I)\rangle \from \left(\prod_{i\in I} A_i\right) \sequoid C \to \prod_{i\in I}(A_i\sequoid C) \\
    \dist_{A,0}\from 1\sequoid A\to 1
  \end{gather*}
  are isomorphisms.
\end{definition}
\begin{definition}
  Let $\C=(\C,\C_s,J,\wk)$ be a sequoidal category.  We say that $\C$ is an \emph{inclusive sequoidal category} if $\C_s$ is a full-on-objects subcategory of $\C$ containing all isomorphisms and finite products of $\C$, and the morphisms $\wk_{A,B}$ and $J$ is the inclusion functor.

%  If $\C$ is an inclusive sequoidal category, we say that $\C$ is \emph{Cartesian} if $\C_s$ has all products and these are preserved by $J$.
 We say that $\C$ is \emph{decomposable} if for any $A$ and $B$, the tensor product $A \tensor B$ is a cartesian product of $A \sequoid B$ and $B \sequoid A$, with projections $\wk_{A,B}:A \tensor B \rightarrow A \sequoid B$ and  $\comp{\wk_{A,B}}{\sym_{A,B}}:A \tensor B \rightarrow B \sequoid A$, and if the product $A \times B$ exists, then $(A \times B) \sequoid C$ is the product of $A \sequoid C$ and $B \sequoid C$, with projections $\pr_l \sequoid \id_C,\pr_r \sequoid \id_C$. 
So if $\C$ has all finite products, 
the natural transformations
  \begin{gather*}
    \dec_{A,B} = \langle \wk_{A,B}, \comp{\wk_{A,B}}{\sym_{A,B}}\rangle\from A\tensor B\to (A\sequoid B)\times (B\sequoid A) \\
   % \dec^0 \from I \to 1 \\
    \dist_{A,B,C} = \langle \pr_1\sequoid \id_C,\pr_2\sequoid\id_C\rangle\from (A\times B)\sequoid C\to (A\sequoid C)\times (B\sequoid C) \\
    \dist_{A,0}\from 1\sequoid A\to 1
  \end{gather*}
  are isomorphisms.
\end{definition}

We shall also use the fact that our category of games is \emph{affine} - so the natural transformation $\dec^0\from I \to 1$ is an isomorphism.  This is not strictly necessary, and techniques are available to obtain similar results in the non-affine case (see \cite{jimNonAffine}, for example), but it will greatly simplify our presentation.

The games model satisfies an analog of monoidal closedness:

@@

  The sequoidal exponential $\oc A$, together with the obvious copycat strategy $\alpha_A\from\oc A\to A\sequoid\oc A$, is the final coalgebra for the endofunctor $A\sequoid\blank$.  In other words, if $B$ is a negative game and $\sigma\from B\to A\sequoid B$ is a morphism then there is a unique morphism $\fcoal{\sigma}\from B\to \oc A$ such that the following diagram commutes:
  \[
    \begin{tikzcd}
      B \arrow[r, "\sigma"] \arrow[d, "\fcoal{\sigma}"']
        & A \sequoid B \arrow[d, "\id_A\sequoid\fcoal{\sigma}"] \\
      \oc A \arrow[r, "\alpha"']
        & A \sequoid \oc A
    \end{tikzcd}
    \]
  We call $\fcoal{\sigma}$ the \emph{anamorphism} of $\sigma$.
\end{proposition}
\begin{proof}
  See \cite{martinsthesis}.  We shall shortly give a proof in the more general case.
\end{proof}

We shall use the following standard pieces of theory about final coalgebras.  
\begin{description}
  \item[Lambek's Lemma] $\alpha_A$ is an isomorphism, with inverse given by the anamorphism of the map $\id\sequoid\alpha_A\from A\sequoid\oc A\to A \sequoid (A \sequoid \oc A)$ \cite{Lambek}.  In particular, $\alpha_A$ is a morphism in $\G_s$.  In the general case, we deduce that $\alpha_A$ is a morphism in $\C_s$.  
  \item[Final Sequence] We may construct a transfinite sequence of objects and morphisms between them (in other words, a functor $\Ord\to\G$, where $\Ord$ is the category of ordinals and prefix inclusions:
    \[
      \begin{tikzcd}
        1
          & A \arrow[l, "*"']
            & A \sequoid A \arrow[l, "\id\sequoid *"']
              & A \sequoid (A \sequoid A) \arrow[l, "\id\sequoid(\id\sequoid*)"']
                & \cdots \arrow[l] \\
        A^{\sequoid\omega} \arrow[u] \arrow[ur] \arrow[urr] \arrow[urrr] \arrow[phantom, urrrr, "\cdots" xshift=0.6em]
          & A \sequoid A^{\sequoid\omega} \arrow[l]
            & A \sequoid (A \sequoid A^{\sequoid\omega}) \arrow[l]
              & A \sequoid (A \sequoid (A \sequoid A^{\sequoid\omega})) \arrow[l]
                & \cdots \arrow[l] \\[-24pt]
        \vdots
          & \vdots
            & \vdots
              & \vdots
                &
      \end{tikzcd}
      \]
    Here, $A^{\sequoid\omega}$ is the categorical limit of the top line and the morphisms into the top line are the morphisms in the limiting cone.  The morphism from $A\sequoid A^{\sequoid\omega}$ to $A^{\sequoid\omega}$ is constructed as follows: apply the functor $A\sequoid\blank$ to the limiting cone from $A^{\sequoid\omega}$, giving us a cone from $A\sequoid A^{\sequoid\omega}$ over the top line.  The desired morphism is now the unique morphism $A\sequoid A^{\sequoid\omega}\to A^{\sequoid\omega}$ that induces a morphism of cones.  We may continue in this way to build up $A^{\sequoid\gamma}$ for all ordinals $\gamma$, where $A^{\sequoid(\gamma+1)}=A\sequoid A^{\sequoid\gamma}$ and $A^{\sequoid\mu}$ is the limit over the $A^{\sequoid\gamma}$ for $\gamma<\mu$ if $\mu$ is a limit ordinal.

    If the morphism $j_{\gamma}^{\gamma+1}\from A\sequoid A^{\sequoid\gamma}\to A^{\sequoid\gamma}$ is an isomorphism for some $\gamma$, then we say that the final sequence \emph{stabilizes at $\gamma$}.  In this case, ${j_{\gamma}^{\gamma+1}}\inv\from A^{\sequoid\gamma}\to A\sequoid A^{\sequoid\gamma}$ is the final coalgebra for the functor $A\sequoid\blank$ \cite{finalseq}.
